---
title: "Snapshots and bookmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Snapshots and bookmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Shiny's bookmarking (`enableBookmarking()`) and shinysnap both let users
return to a saved state. This vignette helps you choose between them,
use them together, or switch an app from bookmarks to snapshot files.

## What each one does

| | bookmarking | shinysnap |
|---|---|---|
| where state is saved | a URL (`"url"`) or a server directory (`"server"`) | a file the user downloads and uploads |
| how it restores | opens the URL in a new session | restores the current session without reloading |
| dynamic inputs | receive saved values when created, through `restoreInput()` | receive saved values when created or when they appear in the browser |
| setup | a UI function and `enableBookmarking()` | no required UI changes |
| file contents | JSON encoded in a URL, or `.rds` files | JSON you can read in a text editor |
| support for app changes | no migration helpers | app name and version, `validate` and `migrate` hooks, and a restore report |

Use bookmarking when you want to send a colleague a link to what you're
looking at. Use shinysnap when you want to save your work to a file and
return to it later, perhaps after the app has been updated.

## Using both

You can add shinysnap to an app that already uses bookmarking:

- Ids excluded with `setBookmarkExclude()` are excluded from snapshots too.
- Values that Shiny marks as unserializable, such as passwords, are never
  saved. This includes values excluded by a serializer registered with
  `setSerializer()`.
- During a restore, shinysnap makes the snapshot's values available to
  `restoreInput()`, which supplies initial values to new inputs. It
  restores the previous settings when the restore finishes.
- shinysnap's own internal inputs are marked unserializable, so they never
  show up in a bookmark URL.

The callbacks, or *hooks*, work similarly in both packages.
`onBookmark(function(state) ...)` and `snap_on_save()` write to
`state$values`. `onRestore()` and `snap_on_restore()` read the saved values.
If you use `reactiveValues`, `snap_track()` can save and restore them for you.

## Turning a snapshot into a bookmark

Use `snap_as_bookmark_url()` to turn a snapshot into a link. The app must
use `enableBookmarking("url")` and a UI function to open it:

```{r}
library(shinysnap)
snap <- list(
  inputs = list(n = 100L, model = "complex", weights = c(0.5, 0.75)),
  values = list(note = "baseline")
)
snap_as_bookmark_url(snap, base_url = "https://example.org/app/")
```

Inside a server function, pass `session` instead of `base_url` to use the
app's current address. The result uses the same `_inputs_` and `_values_`
keys and encoding as a URL created by `session$doBookmark()`.

## Migrating an app

1. Remove `enableBookmarking()` and `bookmarkButton()`. Add
   `snap_download_button()` and `snap_file_input()` to the UI, and
   `snap_download_handler()` and `snap_file_restore()` to the server
   function. The UI no longer needs to be a function.
2. Replace `onBookmark()` hooks with `snap_on_save()`, and `onRestore()`
   hooks with `snap_track()` for `reactiveValues` (they are written back for
   you) or `snap_on_restore()` for anything else.
3. Keep `setBookmarkExclude()` calls; add `snap_enable(exclude = ...)`
   patterns for ids that should never be saved.
4. Give the app a name and a version with `snap_enable()`, and write a
   `migrate` hook the first time a saved value changes meaning.
5. In `shiny::testServer()` tests, use `snap_as_test_inputs()` to prepare
   the saved inputs for a `session$setInputs()` call.
